Elaboration with First-Class Implicit Function Types

### 2020-01-23 paper.tex
stuff

%% \usepackage{fontspec}
%% \fontspec[
%%   Path = fonts/,
%%   BoldFont = *-Bold,
%%   ItalicFont = *-Oblique,
%%   BoldItalicFont = *-BoldOblique]
%%   {DejaVuSansMono}
\setmonofont[Scale=0.8]{DejaVuSansMono}

@@

\textbf{Implicit function types} are a common mechanism for conveying to the
compiler that particular arguments should be inferred by default. In Agda and
Coq, one can use bracketed function domains for this purpose:
\begin{verbatim}
          -- Agda                       (* Coq *)
          id : {A : Set} → A → A        Definition id {A:Type}(x : A) := x.
          id x = x
\end{verbatim}
In GHC, one can use \texttt{forall} to define implicit function
types\footnote{This notation requires language extensions
  \texttt{KindSignatures} and \texttt{RankNTypes}; one could also write the type
  \texttt{a -> a} and GHC would silently insert the quantification.}
\begin{verbatim}
                         id :: forall (a :: *). a -> a
                         id x = x
\end{verbatim}

@@

In Coq, the core language does not have an actual implicit function
type, instead, implicitness is tied to particular \emph{names}, and while we can
write \texttt{List (forall \{A : Set\}, A -> A)}, the brackets here are simply
ignored by Coq. For illustration, Coq accepts the following definition:
\begin{verbatim}
      Definition poly : forall (f : forall {A : Type}, A -> A), bool * nat :=
        fun f => (f bool true, f nat 0).
\end{verbatim}
This is a higher-rank polymorphic function which returns a pair. Note that
\texttt{f} is applied to two arguments, because the implicitness in
\texttt{forall \{A : Type\}, A -> A} is silently dropped.

@@

We prefer to avoid the term ``impredicative inference'' in order to avoid
confusion with impredicativity in type theory, which is an orthogonal notion. In
type theory, impredicativity is a property of a universe, i.e.\ closure of a
universe under arbitrary products. In the type inference literature,
impredicativity means the ability to instantiate type variables and
metavariables to polymorphic types. In particular, we have that
\begin{itemize}
  \item Agda is type-theory-predicative, but implements type-inference-impredicative elaboration.
  \item Coq has type-theory-impredicative $\mathsf{Prop}$ universe, but
    implements type-inference-predicative elaboration.
  \item GHC is type-theory-impredicative with \texttt{RankNTypes} but without
    \texttt{ImpredicativeTypes}, as we have \texttt{(forall a. a -> a) :: *}.

\end{itemize}

>>>>

%% \usepackage{fontspec}
%% \fontspec[
%%   Path = fonts/,
%%   BoldFont = *-Bold,
%%   ItalicFont = *-Oblique,
%%   BoldItalicFont = *-BoldOblique]
%%   {DejaVuSansMono}
\setmonofont[Scale=0.8]{DejaVuSansMono}
\usepackage{fancyvrb}
\usepackage{xcolor}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\newcommand{\red}[1]{\textcolor{red}{#1}}

@@

\textbf{Implicit function types} are a common mechanism for conveying to the
compiler that particular arguments should be inferred by default. In Agda and
Coq, one can use bracketed function domains for this purpose:
\begin{Verbatim}[commandchars=\\\{\}]
          \red{-- Agda}                     \red{(* Coq *)}
          \blue{id} : {A : \blue{Set}} → A → A        \blue{Definition} id \{A:Type\}(x : A) := x.
          id x = x
\end{Verbatim}
In GHC, one can use \texttt{forall} to define implicit function
types\footnote{This notation requires language extensions
  \texttt{KindSignatures} and \texttt{RankNTypes}; one could also write the type
  \texttt{a -> a} and GHC would silently insert the quantification.}
\begin{verbatim}
                         id :: forall (a :: *). a -> a
                         id x = x
\end{verbatim}

@@

In Coq, the core language does not have an actual implicit function type,
instead, implicitness is tied to particular \emph{names}, and while we can write
\texttt{List (forall \{A : Set\}, A -> A)}, the brackets here are simply ignored
by Coq and we get a plain function a type. For illustration, Coq accepts the
following definition:
\begin{verbatim}
      Definition poly : forall (f : forall {A : Type}, A -> A), bool * nat :=
        fun f => (f bool true, f nat 0).
\end{verbatim}
This is a higher-rank polymorphic function which returns a pair. Note that
\texttt{f} is applied to two arguments, because the implicitness in
\texttt{forall \{A : Type\}, A -> A} is silently dropped.

@@

We prefer to avoid the term ``impredicative inference'' in order to avoid
confusion with impredicativity in type theory, which is an orthogonal notion. In
type theory, impredicativity is a property of a universe, i.e.\ closure of a
universe under arbitrary products. In the type inference literature,
impredicativity means the ability to instantiate type variables and
metavariables to polymorphic types. In particular, we have that
\begin{itemize}
  \item Agda is type-theory-predicative, but implements type-inference-impredicative elaboration.
  \item Coq has type-theory-impredicative $\mathsf{Prop}$ universe, but
    implements type-inference-predicative elaboration.
  \item GHC is type-theory-impredicative with \texttt{RankNTypes} but without
    \texttt{ImpredicativeTypes}, as we have \texttt{(forall a. a -> a) :: *}.
\end{itemize}
